---
title: "GROUP 4 - DAMO 520 - Marketing Analytics - Graded Assignment 2"
author: "Tayo Noah Abe, Nihal Vellaramkallingal Sulaiman, Olamide Ibukun"
date: "2025-08-08"
output: html_document
---

# STEP Zero — Set up
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# install.packages(c("readxl","dplyr","ggplot2","stringr","tidyr","lmtest","car","broom","readr"))

library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(lmtest)  # bptest
library(car)     # vif
library(broom)   # tidy model summaries
library(readr)   # write_csv
library(ggplot2)
library(viridisLite)   # palettes
library(viridis)       # scale_*_viridis* helpers
library(tibble)
```


# Step 1
We read the file without skipping any rows so we can use the first data row (full question text) to reliably find the right columns. We then drop that label row.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Step 1 – Load data & capture the label row
df_raw <- read_excel("Purchasing_data.xlsx", sheet = "Sheet1", col_names = TRUE)

# First row contains full survey questions / labels
labels <- as.character(df_raw[1, ])
purchasing_clean   <- df_raw[-1, ]

dim(df_raw); dim(purchasing_clean)  # raw vs. after dropping label row

```
```{r}
# Look at first rows and column names
head(purchasing_clean)
colnames(purchasing_clean)

```

# Step 2 – Robust column finders (by label text)
These patterns helps in matching question text in the label row, so it works even if coded names (Q01_1, D11, etc.) move.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# --- STEP 2 (Robust column matching + fallbacks + sanity table) ---
library(stringr); library(tibble); library(dplyr)

# Expect: labels = as.character(df_raw[1, ]), df_raw[-1, ] is the data body
stopifnot(exists("labels"), exists("df_raw"))

# Never use the name `data` (it masks base::data)
df <- df_raw[-1, , drop = FALSE]
labels <- as.character(labels)

# Guards
stopifnot(is.data.frame(df))
stopifnot(length(labels) == ncol(df))

find_by_label <- function(pat, labels, ignore_case = TRUE) {
  hit <- which(str_detect(labels, regex(pat, ignore_case = ignore_case)))
  if (length(hit) == 0) return(NA_integer_)
  hit[1]
}

# Regex patterns (broad but targeted)
pat_cc_pct   <- "percent(age)?\\s+of\\s+your\\s+purchases.*past\\s*30\\s*days.*credit\\s*card"
pat_numcards <- "total\\s+number\\s+of\\s+credit\\s+cards\\s+you\\s+currently\\s+own"
pat_yob      <- "\\byear\\s+of\\s+birth\\b"
pat_income   <- "approximate\\s+total\\s+personal\\s+income|considering\\s+all\\s+sources\\s+of\\s+income"
pat_savings  <- "about\\s+how\\s+much.*savings.*emergency\\s+expenses"
pat_prime    <- "are\\s+you\\s+currently\\s+subscribed.*amazon\\s*prime"

ix_cc       <- find_by_label(pat_cc_pct, labels)
ix_cards    <- find_by_label(pat_numcards, labels)
ix_yob      <- find_by_label(pat_yob, labels)
ix_income   <- find_by_label(pat_income, labels)
ix_savings  <- find_by_label(pat_savings, labels)
ix_prime    <- find_by_label(pat_prime, labels)

# Fallbacks (1-based)
if (is.na(ix_cc))      ix_cc      <- 21
if (is.na(ix_cards))   ix_cards   <- 18
if (is.na(ix_yob))     ix_yob     <- 80
if (is.na(ix_income))  ix_income  <- 84
if (is.na(ix_savings)) ix_savings <- 89
if (is.na(ix_prime))   ix_prime   <- 77

# Bounds check helper
chk_idx <- function(i, n) {
  if (!is.numeric(i) || length(i) != 1 || is.na(i) || i < 1 || i > n)
    stop("Column index out of range: ", i, " (ncol=", n, ")")
  i
}
ix_cc      <- chk_idx(ix_cc,      ncol(df))
ix_cards   <- chk_idx(ix_cards,   ncol(df))
ix_yob     <- chk_idx(ix_yob,     ncol(df))
ix_income  <- chk_idx(ix_income,  ncol(df))
ix_savings <- chk_idx(ix_savings, ncol(df))
ix_prime   <- chk_idx(ix_prime,   ncol(df))

# Sanity table
matched <- tibble::tibble(
  role    = c("DV: cc_pct","num_cards","year_of_birth","income (text)","savings (text)","Amazon Prime"),
  index   = c(ix_cc, ix_cards, ix_yob, ix_income, ix_savings, ix_prime),
  colname = c(names(df)[ix_cc], names(df)[ix_cards], names(df)[ix_yob],
              names(df)[ix_income], names(df)[ix_savings], names(df)[ix_prime]),
  label_snippet = substr(labels[c(ix_cc, ix_cards, ix_yob, ix_income, ix_savings, ix_prime)], 1, 75)
)
print(matched)

# Quick peek at the first few non-NA values for each matched column
peek_vals <- function(v) {
  x <- as.character(v)
  x <- x[!is.na(x) & nzchar(x)]
  head(x, 3)
}

cat("\nSample values (first 3 non-missing):\n")
cat("  cc_pct   :", paste(peek_vals(df[[ix_cc]]), collapse = " | "), "\n")
cat("  num_cards:", paste(peek_vals(df[[ix_cards]]), collapse = " | "), "\n")
cat("  yob      :", paste(peek_vals(df[[ix_yob]]), collapse = " | "), "\n")
cat("  income   :", paste(peek_vals(df[[ix_income]]), collapse = " | "), "\n")
cat("  savings  :", paste(peek_vals(df[[ix_savings]]), collapse = " | "), "\n")
cat("  prime    :", paste(peek_vals(df[[ix_prime]]), collapse = " | "), "\n")

```

# Step 3 – Clean & engineer variables
in this step three, the code perform the following:
Convert income/savings brackets → numeric midpoints
Create Age = 2025 − Year of birth
Make Prime subscription binary (0/1)
Convert DV to numeric 0–100
```{r, echo=TRUE, message=FALSE, warning=FALSE}

to_midpoint <- function(x) {
  x <- as.character(x)
  if (length(x) == 0) return(numeric(0))
  nums <- stringr::str_extract_all(x, "[0-9][0-9,]*")
  out <- mapply(function(txt, vec) {
    if (length(vec) == 0) return(NA_real_)
    vals <- as.numeric(gsub(",", "", vec))
    txtl <- tolower(txt)
    if (length(vals) >= 2) {
      mean(vals[1:2])                             # "Between $25,000 and $49,999"
    } else if (grepl("less than", txtl)) {
      vals[1] / 2                                  # "Less than $1,000"
    } else if (grepl("more than|over|at least|\\+$", txtl)) {
      vals[1]                                      # "More than $100,000"
    } else if (grepl("no savings|none|i have no savings", txtl)) {
      0
    } else {
      vals[1]
    }
  }, x, nums, USE.NAMES = FALSE)
  as.numeric(out)
}

num_safely <- function(x) as.numeric(gsub("[^0-9.\\-]", "", as.character(x)))

# Require DV to be found
stopifnot(!is.na(ix_cc))

# Small helper: pull a column by its *name* inferred from index
get_col_by_ix <- function(df, ix) {
  if (is.na(ix)) return(rep(NA_character_, nrow(df)))
  df[[ names(df)[ix] ]]
}

# --- Build raw fields from your matched indices (by *name*, not number) ---
raw_df <- tibble::tibble(
  cc_pct_raw   = get_col_by_ix(purchasing_clean, ix_cc),
  cards_raw    = get_col_by_ix(purchasing_clean, ix_cards),
  yob_raw      = get_col_by_ix(purchasing_clean, ix_yob),
  income_txt   = get_col_by_ix(purchasing_clean, ix_income),
  savings_txt  = get_col_by_ix(purchasing_clean, ix_savings),
  prime_raw    = get_col_by_ix(purchasing_clean, ix_prime)
)

# --- Create the analytic dataset used in modeling/plots ---
analytic <- raw_df %>%
  dplyr::transmute(
    cc_pct      = num_safely(cc_pct_raw),                 # DV: 0–100
    num_cards   = num_safely(cards_raw),
    yob         = num_safely(yob_raw),
    age         = dplyr::if_else(!is.na(yob), 2025 - yob, as.numeric(NA)),
    income_mid  = to_midpoint(income_txt),
    savings_mid = to_midpoint(savings_txt),
    prime_sub   = dplyr::case_when(
      is.na(prime_raw) ~ as.numeric(NA),
      stringr::str_detect(tolower(as.character(prime_raw)), "yes|selected|amazon|1|true") ~ 1,
      TRUE ~ 0
    )
  )

# --- Replace num_cards parsing with a robust textual parser ---
if (exists("ix_cards") && !is.na(ix_cards)) {
  # Use the column you already extracted (avoid the name 'data' entirely)
  stopifnot(exists("raw_df"))
  cards_txt <- as.character(raw_df$cards_raw)  # or: as.character(purchasing_clean[[ix_cards]])

  # Normalize
  cards_low <- tolower(trimws(cards_txt))

  # 1) Extract first digit if present (e.g., "2 cards", "3+")
  cards_num <- suppressWarnings(as.numeric(stringr::str_extract(cards_low, "\\d+")))

  # 2) Handle common words
  words_map <- c(
    "none" = 0, "no" = 0, "zero" = 0,
    "one" = 1, "two" = 2, "three" = 3, "four" = 4,
    "five" = 5, "six" = 6, "seven" = 7, "eight" = 8, "nine" = 9
  )

  need_word <- is.na(cards_num) & nzchar(cards_low)
  if (any(need_word)) {
    word_only <- gsub("[^a-z]+", " ", cards_low[need_word])
    word_only <- trimws(word_only)
    wnum <- rep(NA_real_, sum(need_word))
    for (w in names(words_map)) {
      hit <- grepl(paste0("\\b", w, "\\b"), word_only)
      wnum[hit] <- words_map[[w]]
    }
    cards_num[need_word] <- wnum
  }

  # 3) Treat "don't have", "not applicable" etc. as 0
  none_like <- grepl("don'?t.*have|do not.*have|not applicable|n/?a|nil|no card", cards_low)
  cards_num[is.na(cards_num) & none_like] <- 0

  # 4) Assign back into analytic
  analytic$num_cards <- cards_num
} else {
  analytic$num_cards <- NA_real_
}


# Same summary line you had before
summary(dplyr::select(analytic, cc_pct, num_cards, age, income_mid, savings_mid, prime_sub))

```

# Handling of missing values per model
We handled missing value by dropping missing rows per model (not globally), so that each model uses the largest valid of sample:
Model 1 requires cc_pct and num_cards
Multiple/Nonlinear require cc_pct, age, num_cards, income_mid, savings_mid, prime_sub
```{r, echo=TRUE, message=FALSE, warning=FALSE}
m1_df <- analytic %>%
  select(cc_pct, num_cards) %>%
  filter(!is.na(cc_pct), !is.na(num_cards))

m2_df <- analytic %>%
  select(cc_pct, age, num_cards, income_mid, savings_mid, prime_sub) %>%
  filter(complete.cases(.))

tibble::tibble(
  frame = c("Model 1 (simple)", "Model 2/3 (multi)"),
  n = c(nrow(m1_df), nrow(m2_df))
)
```

#Diagnostic
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# === DIAGNOSTIC: what do we actually have after Step 3? ===
diag_counts <- sapply(dplyr::select(
  analytic, cc_pct, num_cards, age, income_mid, savings_mid, prime_sub
), function(x) sum(!is.na(x)))

diag_counts  # look here: any zeros?

# Peek first 10 non-missing of each (helps confirm parsing)
peek <- function(v) head(v[!is.na(v)], 10)
list(
  cc_pct     = peek(analytic$cc_pct),
  num_cards  = peek(analytic$num_cards),
  age        = peek(analytic$age),
  income_mid = peek(analytic$income_mid),
  savings_mid= peek(analytic$savings_mid),
  prime_sub  = peek(analytic$prime_sub)
)

```


#Step 4 – Exploratory analysis (plots)
These will help us to support our variable justification and model choice.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Recoding cc_pct to 0–100 scale
m1_df <- m1_df %>%
  mutate(cc_pct_clean = case_when(
    cc_pct == 20 ~ 20,
    cc_pct == 2040 ~ 30,
    cc_pct == 4060 ~ 50,
    cc_pct == 6080 ~ 70,
    cc_pct == 80100 ~ 90,
    TRUE ~ as.numeric(cc_pct)  # fallback in case
  ))

# DV vs Number of Credit Cards (using cleaned cc_pct)
if (nrow(m1_df) > 0) {
  ggplot(m1_df, aes(x = num_cards, y = cc_pct_clean, color = num_cards)) +
    geom_point(alpha = 0.6, size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    labs(title = "DV vs Number of Credit Cards",
         x = "Number of credit cards",
         y = "Credit Card Purchase % (past 30 days)") +
    theme_classic(base_size = 14) +
    scale_color_viridis_c()
  
  
}


```


```{r, echo=TRUE, message=FALSE, warning=FALSE}

# Recoding cc_pct to 0–100 scale for m2_df
m2_df <- m2_df %>%
  mutate(cc_pct_clean = case_when(
    cc_pct == 20 ~ 20,
    cc_pct == 2040 ~ 30,
    cc_pct == 4060 ~ 50,
    cc_pct == 6080 ~ 70,
    cc_pct == 80100 ~ 90,
    TRUE ~ as.numeric(cc_pct)  # fallback in case
  ))

# DV vs Age (using cleaned cc_pct)
if (nrow(m2_df) > 0) {
  ggplot(m2_df, aes(x = age, y = cc_pct_clean, color = age)) +
    geom_point(alpha = 0.6, size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    labs(title = "DV vs Age",
         x = "Age (years)",
         y = "Credit Card Purchase % (past 30 days)") +
    theme_classic(base_size = 14) +
    scale_color_viridis_c(name = "Age")
  
  
}



```

# DV vs Number of Credit Cards
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# DV vs Income (Midpoint) --> using cleaned cc_pct

if (nrow(m2_df) > 0) {
  ggplot(m2_df, aes(x = income_mid, y = cc_pct_clean, color = savings_mid)) +
    geom_point(alpha = 0.6, size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    labs(title = "DV vs Income (midpoint)",
         x = "Income (approx. midpoint, $)",
         y = "Credit Card Purchase % (past 30 days)") +
    theme_classic(base_size = 14) +
    scale_color_viridis_c(name = "Savings ($)")
  
  
}

```


```{r, echo=TRUE, message=FALSE, warning=FALSE}
  # Quick correlation look
  cor_mat <- round(cor(m2_df, use = "pairwise.complete.obs"), 2)
  cor_mat
```

# Step 5 Model 1: Simple linear regression
```{r, echo=TRUE, message=FALSE, warning=FALSE}
m1 <- if (nrow(m1_df) > 1) lm(cc_pct ~ num_cards, data = m1_df) else NULL
if (!is.null(m1)) {
  summary(m1)
}

```
# Step 6 - Model 2: Multiple linear regression

```{r, echo=TRUE, message=FALSE, warning=FALSE}

m2_linear <- if (nrow(m2_df) > 1) lm(cc_pct ~ age + num_cards + income_mid + savings_mid + prime_sub, data = m2_df) else NULL
if (!is.null(m2_linear)) {
  summary(m2_linear)
}

```
# Step 7 – Model 3: Nonlinear regression
Quadratic age (poly(age, 2, raw=TRUE))
Log transform for skewed savings (log1p(savings_mid))

```{r, echo=TRUE, message=FALSE, warning=FALSE}
m3_nonlinear <- if (nrow(m2_df) > 1)
  lm(cc_pct ~ poly(age, 2, raw = TRUE) + num_cards + income_mid + log1p(savings_mid) + prime_sub,
     data = m2_df) else NULL
if (!is.null(m3_nonlinear)) {
  summary(m3_nonlinear)
}
```

# Step 8 – Compare models (R² / Adjusted R² / RMSE / MAE)
This table helps us in picking the best model (usually by Adjusted R² + clean diagnostics).
```{r, echo=TRUE, message=FALSE, warning=FALSE}
metric_row <- function(name, fit) {
  if (is.null(fit)) return(NULL)
  tibble::tibble(
    Model  = name,
    R2     = summary(fit)$r.squared,
    Adj_R2 = summary(fit)$adj.r.squared,
    RMSE   = sqrt(mean(residuals(fit)^2)),
    MAE    = mean(abs(residuals(fit)))
  )
}

comp <- dplyr::bind_rows(
  metric_row("M1: cc_pct ~ num_cards", m1),
  metric_row("M2: Linear", m2_linear),
  metric_row("M3: Nonlinear (age^2 + log savings)", m3_nonlinear)
)

comp
```

#Step 9 – Diagnostics (assumptions)
Linearity / Normality / Scale-Location / Cook’s: plot(model)
Homoscedasticity: bptest(model)
Multicollinearity: vif(model) (multiple models only)
Outliers: Cook’s distance count
```{r, echo=TRUE, message=FALSE, warning=FALSE}
diag_plots <- function(fit, title = "") {
  if (is.null(fit)) return(invisible(NULL))
  par(mfrow = c(2,2))
  plot(fit, main = title)  # Residuals vs Fitted, Q-Q, Scale-Location, Cook's
  par(mfrow = c(1,1))

  cat("\nBreusch-Pagan (homoscedasticity):\n")
  print(lmtest::bptest(fit))

  if (length(coef(fit)) > 2) {
    cat("\nVIF (multicollinearity):\n")
    print(tryCatch(car::vif(fit), error = function(e) "VIF not applicable"))
  }

  cd <- cooks.distance(fit)
  thr <- 4/length(cd)
  cat("\nInfluential obs (Cook's D >", round(thr,4), "):", sum(cd > thr), "\n")
}

cat("\n--- Diagnostics: Model 2 (Linear) ---\n")
diag_plots(m2_linear, "M2 Linear")

cat("\n--- Diagnostics: Model 3 (Nonlinear) ---\n")
diag_plots(m3_nonlinear, "M3 Nonlinear")
```


```{r, echo=TRUE, message=FALSE, warning=FALSE}
summary(lm(cc_pct ~ num_cards, data = m1_df))
summary(lm(cc_pct ~ age, data = m2_df))
summary(lm(cc_pct ~ income_mid, data = m2_df))
```

# Step 10 – Save cleaned dataset & models (for submission)
```{r, echo=TRUE, message=FALSE, warning=FALSE}
write_csv(analytic, "assignment2_cleaned.csv")

saveRDS(list(
  m1 = m1,
  m2_linear = m2_linear,
  m3_nonlinear = m3_nonlinear,
  comparison = comp
), file = "assignment2_models.rds")

"Saved: assignment2_cleaned.csv and assignment2_models.rds"

```
